Sparse Matrix Vector Multiplication with Multiple vectors and Multiple Matrices on the MIC Architecture

### 2013-10-16 ieee_paper.tex
citations

\section{Introduction}
\cite{Bell08, Vuduc05, Nishtala07, Stock12-TACC, Cuthill69, cramer2012openmp, Buluc2009_SPAA, Buluc11, Im01, Mellor-Crummey04, Nishtala07, Saad94sparskit, Williams07}
\cite{Temam:1992:CBS:147877.148091}
\cite{DBLP:journals/ijhpca/ShantharamCR11,
conf/ppsc/Toledo97, Liu:2013:ESM:2464996.2465013, Molka:2009:MPC:1636712.1637764,%
DBLP:journals/corr/abs-1101-0091, conf/ipps/BulucWOD11, conf/ipps/KreutzerHWFBB12,%
kumar2012accelerating, journals/concurrency/VazquezFG11}

SpMV is an important kernel for lots of stuff. So improving the
performance of SpMV has caputred the interest of many researcher.

The main challenge that is faced to improve good performance for SpMV
is that the operation are conducted using memory location that are
irregular and often unpredictible. That make that the kernels are
mostly memory bound and there is a significant instruction overhead
per flop.

Common improvement techniques such as register blocking, bandiwdth
reduction (matrix reordering), partitioning to fit in cache or TLB
have impacts which are very dependent on the matrix and overall do not
lead to dramatic improvement. Assuming register blocking does not
apply well to the matrix at hand (which is true for most matrices),
there is about 8 bytes of the matrix to move in per non zero (assuming
single precision); each nonzero requires two floating point operations
leading to a flop-to-byte ratio of at most $\frac{1}{4}$. This limits
the obtained performance to at most a quarter of the bandwidth of the
architecture wasting a lot of potentially useful cycles. The commonly
used techniques are mostly designed to reach that bound rather than
overcome it.

@@

To perform our analysis, we focus our attention on the improvement
that we can achieve on the Intel Xeon Phi processor. It follow the
Many Integrated Core (MIC) architecture, which has a significant
memory bandwidth and peak flop throughput thanks to its 512-bit large
SIMD registers. The Xeon Phi processor has been shown to be promising
for sparse linear algebra compared to more classical CPU or GPU
architecture~\cite{}.

@@

For a total of $N$ points, there will be $N$ linear systems to solve,
each of size $n_z \times n_z$. Each linear solve produces a row of the
RBF-FD differentiation matrix $D_{n_z}$, resulting in a $N \times N$
matrix with $n_zN$ nonzero entries. To evaluate the derivative at all
points in the domain, takes the form
$$
  \mathbf{g} = D_{n_z} \mathbf{f}
$$
where $f$ is the source vector of function values and $g$ is the
resulting vector of derivative values. The computation of a single
derivative has been reduced to a SpMV, where each row has $n_z$
nonzeros. Since the sparsity of each row is constant, ELLPACK~\cite{}
becomes the most appropriate matrix compression scheme\cite{}. In
practice, $n_z=32$ in two-dimensional flows and 64 or 100 for
three-dimensional flows. These numbers are similar to what is used in
finite-element codes.

@@

\def\wide{2.5in}
\newfig{figures/rbf_stencils.pdf}{\wide}{RBFFD Stencils. Each node of the stencil is connected to $n_z-1$ stencil nodes in addition to itself. In the figure, node $A$ is connected to $B$, but $B$ is {\em not\/} connected to $A$. Thus adjacency graph of $A$ is not-symmetric. This must be corrected when reducing the bandwidth of $A$ with a Cuthill\-McKee algorithm, which assumes symmetry.}{fig:rbf_stencils}

\newfig{figures/matrix_structure.pdf}{\wide}{Matrix Structure.}{fig:mat_struct}

>>>>

\section{Introduction}
\cite{Stock12-TACC} is all dense. \cite{
  Molka:2009:MPC:1636712.1637764} is about memory modeling not
directly spmv.

\cite{Im01, Saad94sparskit} \cite{DBLP:journals/ijhpca/ShantharamCR11,
  DBLP:journals/corr/abs-1101-0091 }

The multiplication of sparse matrix by a dense vector (SpMV) is an
important kernel in many fields from scientific computing~\cite{} to
graph analysis (recommender systems~\cite{Brin98}, graph
drawing~\cite{Koren05}). So improving the performance of SpMV has
captured the interest of many researcher; including the development of
various implemetation for multi-core
CPUs~\cite{Buluc2009SPAA,Williams07} and
GPUs~\cite{Bell08, conf/ipps/KreutzerHWFBB12,
  journals/concurrency/VazquezFG11,kumar2012accelerating}. The main
challenge that is faced to obtain good performance for SpMV is that
the operation are conducted using memory location that are irregular
and often unpredictible. That make that the kernels are mostly memory
bound and there is a significant instruction overhead per flop.

Common improvement techniques such as register blocking, bandwidth
reduction (matrix reordering~\cite{Cuthill69}), partitioning to fit in
cache or TLB~\cite{Nishtala07,Temam:1992:CBS:147877.148091,conf/ppsc/Toledo97},
unrolling~\cite{Mellor-Crummey04} have impacts which are very
dependent on the matrix and overall do not lead to dramatic
improvement. (The state-of-the-art techniques in OSKI~\cite{Vuduc05}
provide some useful yet limited improvements). Register blocking~\cite{conf/ppsc/Toledo97} does
not apply well to most the matrix at hand (despite it can be applied
with virtually no overhead in all cases thanks to compressed
representation~\cite{Buluc11}). there is about 8 bytes of the matrix to move
in per non zero (assuming single precision); each nonzero requires two
floating point operations leading to a flop-to-byte ratio of at most
$\frac{1}{4}$. This limits the obtained performance to at most a
quarter of the bandwidth of the architecture wasting a lot of
potentially useful cycles. The commonly used techniques are mostly
designed to reach that bound rather than overcome it.

@@

To perform our analysis, we focus our attention on the improvement
that we can achieve on the Intel Xeon Phi processor. It follow the
Many Integrated Core (MIC) architecture, which has a significant
memory bandwidth and peak flop throughput thanks to its 512-bit large
SIMD registers. The Xeon Phi processor has been shown to be promising
for sparse linear algebra compared to more classical CPU or GPU
architecture~\cite{Saule13-ARXIV, Liu:2013:ESM:2464996.2465013, cramer2012openmp}.

@@

For a total of $N$ points, there will be $N$ linear systems to solve,
each of size $n_z \times n_z$. Each linear solve produces a row of the
RBF-FD differentiation matrix $D_{n_z}$, resulting in a $N \times N$
matrix with $n_zN$ nonzero entries. To evaluate the derivative at all
points in the domain, takes the form
$$
  \mathbf{g} = D_{n_z} \mathbf{f}
$$

where $f$ is the source vector of function values and $g$ is the
resulting vector of derivative values. The computation of a single
derivative has been reduced to a SpMV, where each row has $n_z$
nonzeros. Since the sparsity of each row is constant, ELLPACK~\cite{}
becomes the most appropriate matrix compression scheme\cite{}. In
practice, $n_z=32$ in two-dimensional flows and 64 or 100 for
three-dimensional flows. These numbers are similar to what is used in
finite-element codes. Notice that the non zero elements of the matrix
correspond to a relation of nearest neighbor. This relationship is not
necessarily symmetric as shown in Figure~\ref{fig:rbf_stencils} which
means that the matrix sparsity pattern might not be symmetric either.

\begin{figure}[tbh]
  \centering
  \includegraphics[width=\linewidth]{figures/rbf_stencils.pdf}
  \caption{RBFFD Stencils. Each node of the stencil is connected to
    $n_z-1$ stencil nodes in addition to itself. In the figure, node
    $A$ is connected to $B$, but $B$ is {\em not\/} connected to
    $A$. Thus adjacency graph of $A$ is not-symmetric.}
  \label{fig:rbf_stencils}
\end{figure}

@@

\newfig{figures/matrix_structure.pdf}{\linewidth}{Matrix Structure.}{fig:mat_struct}
